\documentclass{tex/tufte-handout}
\input{tex/preamble}

\title{Notes on Macroeconometrics}
\author{Econometrics Free Library Project}
\date{Version~0.1.3, 20 May 2013}

\begin{document}
\maketitle

\bigskip\noindent%
Copyright © 2013, Gray Calhoun.  Permission is granted to copy,
distribute and/or modify this document under the terms of the GNU Free
Documentation License, Version 1.3 or any later version published by
the Free Software Foundation; with no Invariant Sections, no
Front-Cover Texts, and no Back-Cover Texts.  A copy of the license is
included in the Appendix entitled ``GNU Free Documentation License''
and is also available online at
\url{http://www.gnu.org/copyleft/fdl.html}.

This text was produced as part of the Econometrics Free Library
project.  The project's goal is to produce high-quality free (and
open-source) Econometrics textbooks and reference material.  More
information about this project is available at its homepage,
\url{http://www.econometricslibrary.org}, including information on how
to participate.  This text is typeset in \XeLaTeX\ and there is a link
to the source of the document at the project's homepage as well.

\tableofcontents

\section{A quick comment about these notes}
This text started as a set of lecture notes that Gray Calhoun wrote to
himself to teach half of a semester-long Macroeconometrics PhD
elective at Iowa State in 2013.  He (I) had no plans to disseminate
them until the course was over.  As a first step, the notes have been
converted to a \XeLaTeX\ document, but there is a lot of work left
before they would be useful to anyone trying to learn this material.

\include{tex/varma-properties}
\include{tex/varma-estimation}
\include{tex/svar}
\include{tex/stochastic-integrals}
\include{tex/cointegration}
\appendix
\include{LICENSE}

\part*{References}
\addcontentsline{toc}{part}{References}
\bibliography{tex/references}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "textbook"
%%% End: